---
title: "Homwework Chapters 10 + 11"
author: "P Winston Miller"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
library(dagitty)
library(patchwork)
library(ggdag)
library(paletteer)
set_cmdstan_path("/Users/winston/miniconda3/envs/stan/bin/cmdstan")
```

```{r}
data("NWOGrants")
d = NWOGrants
```

```{r}
str(d);head(d)
```

```{r}
dg1 = dagitty("dag{
  gender -> awards
  gender -> discipline
  gender -> applications
  discipline -> applications
  discipline -> awards
  awards -> discipline
  applications -> awards
  }")
ggdag(dg1)
```

```{r}
d
```

# Total Causal Effect

-   probably just gender regressed on application awards

```{r}
d1_params = alist(
  awards ~ dbinom(applications,p),
  logit(p) <- alpha1[gender],
  alpha1[gender] ~ dnorm(0,1.5))
d_in = list(applications = d$applications, gender = ifelse(d$gender == "m",1,2), awards = d$awards)
m1 = ulam(d1_params, data = d_in, log_lik = T, chains = 4, cores = 4)
```

```{r}
summ1 = precis(m1, depth = 2, corr = T)
row.names(summ1) = c("Male","Female")
summ1$or = exp(summ1$mean)
summ1
```

-   odds of applications of being accepted are actually slightly higher for
    females than males
-   lets compute contrasts

```{r}
post1 = extract.samples(m1)
contrast1 = data.frame(male = inv_logit(post1$alpha1[,1]),
                       female = inv_logit(post1$alpha1[,2])) |>
  mutate(contrast = male - female)
ggplot(data = contrast1, aes(x = contrast)) + 
  geom_density(color  = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred") +
  scale_x_continuous(labels = scales::percent, breaks = seq(-1, 1, by = 0.01)) +
  scale_y_continuous(labels = NULL) +
  theme_minimal() +
  labs(x = "Contrast Male-Female \nFunding Rates", y = NULL)
```

-   so males have a little less than 3% higher probability to be awarded a grant

# Direct Causal Effect

-   now need to estimate direct effect
-   I could see 2 different approaches
-   first just fitting two different alphas, 1 for each discipline
-   this seems like it would give the direct effect of gender while controlling
    for discipline
-   and vice versa
-   second would be to create an interaction for each gender discipline combo
-   seems like it would give the effect of gender within discipline

Option 1

```{r}
d11_params = alist(
  awards ~ dbinom(applications,p),
  logit(p) <- alpha1[gender] + alpha2[discipline],
  alpha1[gender] ~ dnorm(0,1.5),
  alpha2[discipline] ~ dnorm(0,1.5)
)
d_in1 = list(discipline = as.integer(d$discipline), applications = d$applications, 
            gender = ifelse(d$gender == "m",1,2), awards = d$awards)
m11 = ulam(d11_params, data = d_in1, log_lik = T, chains = 4, cores = 4)
```

```{r}
summ11 = precis(m11, depth = 2, corr = T)
row.names(summ11) = c(unique(d$gender),unique(d$discipline))
summ11$or = exp(summ11$mean)
```

Option 2

```{r}
d_in2 = list(applications = d$applications, 
             awards = d$awards,
             # gender = d$gender,
             # discipline = d$discipline,
            gender_discipline = as.integer(as.factor(paste0(d$gender, d$discipline))))
d12_params = alist(
  awards ~ dbinom(applications,p),
  logit(p) <- alpha[gender_discipline],
  alpha[gender_discipline] ~ dnorm(-1,1.5)
)
m12 = ulam(d12_params, data = d_in2, log_lik = T, chains = 4, cores = 4)
```

```{r}
summ12 = precis(m12, depth = 2, corr = T)
row.names(summ12) = paste0(d$gender, " - ", d$discipline)
summ12$or = exp(summ12$mean)

summ11;summ12
```

-   so rhat for the second model is better
-   ess_bulk is way better!
-   second model allows me to compare male-female within discipline
-   first model allows me to compare male-female overall adjusting for
    discipline and discipline overall adjusting for male-female
-   this is probably due to the limited amount of data, and difference between 1
    and 2 parameters
-   lets do PSIS

```{r}
compare(m11,m12, func = "PSIS")
```

-   m12 has all the weight, but standard error of the difference is very high
-   since there are 2 pareto K warnings, that means both models have outliers

```{r}
compare(m11,m12, func = "WAIC")
```

-   same story here, lets inspect points

```{r}
paretok = PSIS(m11, pointwise = T)
row.names(paretok) = (paste0(d$gender," - ",d$discipline))
paretok |> arrange(desc(k))
```

-   above 1 is very high, and near 1 is preeetty high I think
    -   this means these exert undue influence on the value
-   so it seems that males in physics and interdisciplinary science are unduly
    influencing model 1
-   lets look at model 2

```{r}
paretok = PSIS(m12, pointwise = T)
row.names(paretok) = (paste0(d$gender," - ",d$discipline))
paretok |> arrange(desc(k))
```

-   all of the values are super high due to the limited data I think
-   lets compute contrasts for this model

```{r}
post12 = extract.samples(m12)
contrast12 = data.frame(
  contrasts = c(
    inv_logit(post12$alpha[,1]) - inv_logit(post12$alpha[,2]),
    inv_logit(post12$alpha[,3]) - inv_logit(post12$alpha[,4]),
    inv_logit(post12$alpha[,5]) - inv_logit(post12$alpha[,6]),
    inv_logit(post12$alpha[,7]) - inv_logit(post12$alpha[,8]),
    inv_logit(post12$alpha[,9]) - inv_logit(post12$alpha[,10]),
    inv_logit(post12$alpha[,11]) - inv_logit(post12$alpha[,12]),
    inv_logit(post12$alpha[,13]) - inv_logit(post12$alpha[,14]),
    inv_logit(post12$alpha[,15]) - inv_logit(post12$alpha[,16]),
    inv_logit(post12$alpha[,17]) - inv_logit(post12$alpha[,18])),
  sciences = c(rep("Chemical Sciences", 2000),
               rep("Physical Sciences", 2000),
               rep("Physics", 2000),
               rep("Humanities", 2000),
               rep("Technical Sciences", 2000),
               rep("Interdisciplinary", 2000),
               rep("Earth&Life Sciences", 2000),
               rep("Social Sciences", 2000),
               rep("Medical Sciences", 2000)))

ggplot(data = contrast12, aes(x = contrasts)) + 
  geom_density(aes(color = sciences)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred") +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  theme_minimal() +
  theme(angle = 45) +
  labs(x = "Contrast Probability of\n Male-Female Funding Rates", y = NULL, 
       color = "Scientific Discipline") +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
  
```

-   as can be seen, different sciences have different numbers of applications &
    differences in funding rate for men and women
-   most obvious is that women are funded at a higher rate then men in earth &
    life sciences, but men are funded at a higher rate in interdisciplinary
    sciences
