---
title: "HW Chapters 11 & 12"
author: "P Winston Miller"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
library(gtools)
library(dagitty)
library(ggdag)
library(patchwork)
library(paletteer)
set_cmdstan_path("/Users/winston/miniconda3/envs/stan/bin/cmdstan")
```

```{r}
data("reedfrogs")
d = reedfrogs
```

# Problem 1

-   simulating priors
-   trying to simulate tank survival probability, the survival being the alpha-j
    parameter
-   this is dependent on alpha_bar and sigma

```{r}
n = 1e4
alpha_bar = rnorm(n,0,1)
sigma = rexp(n,0.5)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

sigma = rexp(n,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

sigma = rexp(n,3)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

sigma = dexp(n,5)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

sigma = dexp(n,10)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)
```

-   decreasing sigma below 1 concentrates probability near the extremes of the
    distribution
-   sigma at 1 basically spreads it out more or less evenly
-   increasing sigma decreases probability concentration at the edges and moves
    it to the center

```{r}
n = 1e4
sigma = rexp(n,1)
alpha_bar = rnorm(n,0.001,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

alpha_bar = rnorm(n,0.5,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

alpha_bar = rnorm(n,0.1,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

alpha_bar = rnorm(n,1,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

alpha_bar = rnorm(n,3,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

alpha_bar = rnorm(n,5,1)
dens(inv_logit(rnorm(n,alpha_bar,sigma)),xlim=c(0,1),adj=0.1,lwd=4,col=2)

```

-   so the smaller it goes, it seems to more evenly spread probability
-   the higher the mean in alpha, the higher probability moves towards 1

# Problem 2
