---
title: "HW-7-8-9"
author: "P Winston Miller"
date: "2026-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
library(dagitty)
library(ggdag)
library(patchwork)
set_cmdstan_path("/Users/winston/miniconda3/envs/stan/bin/cmdstan")
```

# Problem 1

```{r}
d = sim_happiness(seed = 1992, N_years = 1000, max_age = 75, N_births = 20, aom = 18)
```

```{r}
d = d %>% filter(age > 17)
d$married_std = ifelse(d$married == 0, 1, 2)
d$age_std = (d$age - 18)/(75-18)
```

```{r}
precis(d)
```

```{r}
d_in = list(a = d$age_std, m = d$married_std, h = d$happiness)
```

```{r}
hw1 = ulam(alist(
  h ~ dnorm(mu, sigma),
  mu <- alpha[m] + bA*a,
  alpha[m] ~ dnorm(0,1),
  bA ~ dnorm(0,2),
  sigma ~ dexp(1)), 
  data = d_in, 
  chains = 4, cores = 4, iter = 1500, log_lik = T)
```

-   there were a few warnings about divergences but not bad, lets check the
    trace plots

```{r}
traceplot(hw1)
```

-   trace plots look okay

```{r}
precis(hw1, depth = 2)
```

-   rhat looks good
-   ess_bulk also appears okay
-   so this model is fairly certain that age decreases happiness
-   however, married people have a baseline of happiness much higher than
    non-married people
    -   therefore, as they age, they don't get quite so unhappy

```{r}
d_in2 = list(a = d$age_std, h = d$happiness)
hw12 = ulam(alist(
  h ~ dnorm(mu, sigma),
  mu <- alpha + bA*a,
  alpha ~ dnorm(0,1),
  bA ~ dnorm(0,2),
  sigma ~ dexp(1)), 
  data = d_in2, 
  chains = 4, cores = 4, iter = 1500, log_lik = T)
```

```{r}
precis(hw12)
```

-   this model is pretty certain that age has nothing at all to do with
    happiness

```{r}
compare(hw1, hw12, func = "WAIC")
compare(hw1, hw12, func = "PSIS")
```

-   according to information criteria, the model including marriage is a much
    better out-of-bag predictor
-   however, I know from the dag + the simulation that marriage is a collider,
    and when we condition on a collider, we create a statistical association
    that is not appropriate
-   so the association between age and happiness is just a mirage

```{r}
data(foxes)
str(foxes)
```

```{r}
d = data.frame(
  group = as.factor(foxes$group),
  size = scale(foxes$groupsize),
  area = scale(foxes$area),
  food = scale(foxes$avgfood),
  weight = scale(foxes$weight)
)
str(d)
```

```{r}
hw21 = ulam(alist(
  weight ~ dnorm(mu, sigma),
  mu <- alpha[group] + bS*size + bA*area + bF*food,
  alpha[group] ~ dnorm(0,0.2),
  c(bS,bA,bF) ~ dnorm(0,0.5),
  sigma ~ dexp(1)),
  data = d, 
  chains = 4, cores = 4, iter = 2000, log_lik = T)
```

```{r}
traceplot(hw21, pars = c("bF", "bA", "bS", "sigma"))
```

```{r}
precis(hw21)
```

-   above model is just using all the variables to predict the fox weight
-   using the dag below, it seems obvious that food is a mediator between area
    and weight + group size
-   I think that removing area is appropriate based on this
-   using the index approach for group number in the alpha accounts for other
    unobserved factors about each group, potential:
    -   male/female mix
    -   kits
    -   genetic differences
    -   number of predators

```{r}
fox_dag = dagitty("dag{
  area -> food
  food -> gsize
  food -> weight
  gsize -> weight}")
ggdag(fox_dag) + theme_dag_gray()
```

```{r}
hw22 = ulam(alist(
  weight ~ dnorm(mu, sigma),
  mu <- alpha[group] + bS*size + bF*food,
  alpha[group] ~ dnorm(0,0.2),
  c(bS,bF) ~ dnorm(0,0.5),
  sigma ~ dexp(1)),
  data = d, 
  chains = 4, cores = 4, iter = 2000, log_lik = T)
```

-   no divergence warnings which is good

```{r}
traceplot(hw22, pars = c("bF", "bS", "sigma"))
```

-   traceplots look good

```{r}
precis(hw22)
```

-   this doesn't really change the paramaters all that much
-   rhat value is still good
-   lets create model without the indexed intercept

```{r}
hw23 = ulam(alist(
  weight ~ dnorm(mu, sigma),
  mu <- alpha + bS*size + bF*food,
  alpha ~ dnorm(0,0.2),
  c(bS,bF) ~ dnorm(0,0.5),
  sigma ~ dexp(1)),
  data = d, 
  chains = 4, cores = 4, iter = 2000, log_lik = T)
```

-   I expect this one to be the least accurate

```{r}
traceplot(hw23, pars = c("bF", "bS", "sigma"))
```

```{r}
precis(hw23)
```

-   interestingly, removing the indexing did not change the values of bS and bF
    much
-   rhat is good, but ess_bulk is lower
-   lets compare

```{r}
compare(hw21,hw22,hw23)
```

-   interstingly, the highest weight is attributed to hw21, however, the
    standard error of difference is much higher than the actual difference
    between hw22 and hw21, this is true for all differences
-   the model weights are also very similar
-   not to mention the parameter values are similar as well
-   therefore, it is up to me to choose the best model and based on causal dag,
    I choose model hw22
