---
title: "Chapter 9"
author: "P Winston Miller"
date: "2026-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
library(dagitty)
library(ggdag)
library(patchwork)
set_cmdstan_path("/Users/winston/miniconda3/envs/stan/bin/cmdstan")
```

# Markov Chain Monte Carlo (MCMC)

-   basically a way to estimate the posterior distribution through sampling
-   various samplers, Hamiltonian, Gibbs and Metropolis-Hastings are what is covered in the
    book
-   INLA is another sampling method
-   Dick claims Hamiltonian Monte Carlo is the best at time of writing, it is
    also what is built into stan which is the underlying package for brms and
    pymc3
-   INLA uses gaussian random markov fields, which are more limited in model
    estimation, but are much much faster than MH and hamiltonian with a similar accuracy
    according to my brief research
- frankly, I do not understand how these work and don't feel that it is necessary to do so
  as Daniel used to always say, some things just need to be black magic

### Repeating the ruggedness-log_gdp example but now using the MCMC sampler

```{r}
data(rugged)
d = data.frame(
  log_gdp = log(rugged$rgdppc_2000_m),
  ruggedness = rugged$rugged,
  africa = rugged$cont_africa)
d = d[complete.cases(d),]
d$log_gdp_std = d$log_gdp/mean(d$log_gdp)
d$ruggedness_std = d$ruggedness/max(d$ruggedness)
d$africa = ifelse(d$africa == 1, 1, 2)
r_bar = mean(d$ruggedness_std)
d$r_std = d$ruggedness_std - r_bar
d_in = list(africa = d$africa, r_std = d$r_std, log_gdp_std = d$log_gdp_std)
```

```{r}
params_list = alist(
  log_gdp_std ~ dnorm(mu, sigma),
  mu <- alpha[africa] + bR[africa]*r_std,
  alpha[africa] ~ dnorm(1, 0.2),
  bR[africa] ~ dnorm(0,0.25),
  sigma ~ dexp(1)
)
```

quadratic approximation method

```{r}
mq = quap(params_list, data = d_in)
```

mcmc method

-   input for this is trickier
-   requires the input to contain only that which is used for the model, but I
    already did that which is fine

```{r}
mm = ulam(params_list, data = d_in, chains = 4, cores = 4)
```

```{r}
show(mm)
```

-   so we got 2000 samples from all 4 chains which is expected
    -   total samples = 1000
    -   half are warm-up = 500
    -   rest are inference = 500
    -   4 chains = 4\*500 = 2000
-   the formula also looks correct

```{r}
precis(mq, depth = 2)
```

```{r}
precis(mm, depth = 2)
```

-   as expected both results are the same for this simple model
-   rhat is 1 or very close, which indicates convergence of the markov chain so
    the inferences can be "trusted"
-   ess_bulk indicates the number of samples fro the chain. Due to the
    anti-correlation properties of some samples, the total number of effective
    samples can be higher than what was requested
-   I don't entirely understand how the Stan engine works but I can't understand
    everything so I am content to know how to use the diagnostics and move on

Trace Plots

-   these are a visualization of the sampling process of HMC
-   they provide a quick diagnostic look
-   an ideal chain will stay around a central tendency, with fewer extreme
    values
-   specifically look for 3 things
    1.  Stationarity -\> the central tendency or path of the chain remains
        stable
    2.  Good mixing -\> the chain rapidly explores the full posterior region,
        and then continues to explore it (so, bounces up and down a reasonable
        amount)
    3.  Convergence -\> multiple chains show similar behavior
-   He prefers trankplots, which are histograms of all the chains
    -   This is new, I have never used this before but I like it
    -   The big question is whether these are easy to make for other packages..

```{r}
traceplot(mm, chains = c(1,2,3,4))
```

-   these are trace plots for each parameter for all 4 chains
-   the grey area is the warmup, and the white is the sampling
-   these are well behaved chains, they all cover the same space, they have a
    central tendency, and tend to quickly explore the full posterior area

```{r}
trankplot(mm)
```

-   it seems like one chain, and never the same one, will give more extreme
    values for different parameters
-   I find these trank plots more confusing, I think there are better
    diagnostics and the trace plot is good enough for me

Examples of bad chains

-   first lets simulate bad data and a bad model

```{r}
set.seed(11)
y = c(-1,1)
m_bad =  ulam(alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10000),
  sigma ~ dexp(0.0001)), data=list(y=y) ,chains = 3, cores = 3)

```

-   interesting that for a deliberately bad model, only 5% of transitions ended
    in divergence

```{r}
precis(m_bad)
```

-   well this is way off
-   rhat is not 1 or even close to 1 and the effective sample size is a fraction
    of the 1500 it should be
-   mu should be 0 basically with a sd between 1 and 2
-   lets look at the trace plots

```{r}
traceplot(m_bad, chains = 1)
traceplot(m_bad)
```

-   so that shows neither stationary, convergence, nor good mixing
-   I bet I can fix this by adding a sensible prior

```{r}
set.seed(11)
y = c(-1,1)
m_better =  ulam(alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10),
  sigma ~ dexp(1)), data=list(y=y) ,chains = 3, cores = 3)
precis(m_better)
```

-   estimates are much better, effective sample size is still bad

```{r}
traceplot(m_better)
```

-   still not real great, but a little better, there is at least a central
    tendency and some convergence, although the mixing is not great

```{r}

```
