---
title: "Chapter 12"
author: "P Winston Miller"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
library(gtools)
library(dagitty)
library(ggdag)
library(patchwork)
library(paletteer)
set_cmdstan_path("/Users/winston/miniconda3/envs/stan/bin/cmdstan")
```

# Beta Simulation

-   this uses the beta distribution as the outcome variable
    -   has 2 parameters, pbar and theta
    -   pbar is the central tendency, so the average probability for each count
    -   theta is a shape, helping control the width of the distribution

```{r}
curve( dbeta2(x,0.5,0) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,2) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,2.5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,7.5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,10) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,50) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
```

-   so it seems like the higher that theta is, the more peaked the distribution
    around pbar and the thinner the tails
-   0 shows that there are no sampled probabilities at any space
-   at theta = 2, the distribution is basically uniform where all probabilities
    are equally likely
-   greater than 2 is starts to take on that bell curve shape

```{r}
curve( dbeta2(x,0,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.1,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.25,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.5,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,0.75,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
curve( dbeta2(x,1,5) , from=0 , to=1 ,
xlab="probability" , ylab="Density" )
```

-   pbar is a probability, so bound between 0 and 1
-   seems that setting this to 0 or 1 means that there are no samples
    probabilities
-   okay, so changing pbar just changes the location of the peak/tail since it
    is bounded between 0 and 1
-   so the higher/lower it is, the longer the tail and the more dense the
    probability is at any given theta (shape)

# Beta-Binomial Model

-   uses the beta distribution as the outcome
-   so, each binomial count is considered to have its own probability
-   I bet feeding this kind of model aggregated data and row-wise data would
    have a very different effect
-   linear model is bound to pbar so changes in coefficients affect change of
    average probability

```{r}
data("UCBadmit")
d = UCBadmit
d
```

```{r}
d_in = list(admit = d$admit, 
               applications = d$applications, 
               gender = ifelse(d$applicant.gender == "male",1,2))
params = alist(
  admit ~ dbetabinom(size = applications, prob = pbar, theta = theta),
  logit(pbar) <- alpha[gender],
  alpha[gender] ~ dnorm(0,0.5),
  transpars> theta <<- phi + 2.0,
  phi ~ dexp(1))
m1 = ulam(params, data = d_in, chains = 4, cores = 4, log_lik = T)
```

-   tagging theta with "transpars\> theta \<" tells stan to return theta as a
    parameter, as opposed to returning phi
-   I feel like we could just add 2 to phi, but this may be important for more
    complicated situations

```{r}
post = extract.samples(m1)
post$diff_alpha = inv_logit(post$alpha[,1]) - inv_logit(post$alpha[,2])
precis(post, depth = 2)
```

-   so remember we are looking at log-odds for alpha
-   there are 12 pbars because there are 12 rows of data - each a combination of
    department and gender
-   and then I transformed diff alpha to look at the difference in probability
    between men and women
-   so there is basically no difference between male and female admissions,
    which is what we found when accounting for department in the last model,
    when not accounting for department, there was a difference!

```{r}
postcheck(m1)
```

-   so, still not a god model, however, the model does clearly pick up on
    heterogenaity across rows and that can be seen here
    -   the rows remember are department-gender combinations
    -   6 departments, 2 genders = 12 rows
-   empty circles are predictions, bars are 89% HPDI, and blue dots are actuals
-   So, there are clearly 8 of 12 rows where it completely misses actuals
-   I can see why he used this, but, this is not a good model
-   lets look at the posterior probaility lines for each gender

```{r}
gid = 2
# draw posterior mean beta distribution
curve( dbeta2(x,mean(logistic(post$a[,gid])),mean(post$theta)) , from=0 , to=1 ,
ylab="Density" , xlab="probability admit", ylim=c(0,3) , lwd=2 )
# draw 50 beta distributions sampled from posterior
for ( i in 1:50 ) {
p <- logistic( post$a[i,gid] )
theta <- post$theta[i]
curve( dbeta2(x,p,theta) , add=TRUE , col=col.alpha("black",0.2) )
}
mtext( "distribution of female admission rates" )

gid = 1
# draw posterior mean beta distribution
curve( dbeta2(x,mean(logistic(post$a[,gid])),mean(post$theta)) , from=0 , to=1 ,
ylab="Density" , xlab="probability admit", ylim=c(0,3) , lwd=2 )
# draw 50 beta distributions sampled from posterior
for ( i in 1:50 ) {
p <- logistic( post$a[i,gid] )
theta <- post$theta[i]
curve( dbeta2(x,p,theta) , add=TRUE , col=col.alpha("black",0.2) )
}
mtext( "distribution of male admission rates" )
```

-   so the thick black line is the curve corresponding to the posterior mean
-   the others are the 50 most samples combinations of pbar and theta

# Negative Binomial

-   also called gamma-poisson
-   similar in concept to beta-binomial, where each poisson count has its own
    rate parameter = lambda/unit of time or distance or whatever
-   so this uses a gamma distribution to estimate the shape of the distribution
    of poisson rates across all cases
-   the lambda parameter is basically the same as a poisson parameter (including
    offset I think)
-   Lets try to the oceanic tools example and see how it works here

```{r}
data(Kline)
d = Kline
d
```

-   so, only 10 points, very very small

```{r}
d$log_pop_std = scale(log(d$population))
d_in = list(pop = d$log_pop_std,
            contact = ifelse(d$contact == "low",1,2),
            tools = d$total_tools)
m2 = ulam(alist(
  tools ~ dgampois(mu = lambda, scale = phi),
  log(lambda) <- alpha[contact] + beta[contact]*pop,
  alpha[contact] ~ dnorm(3,0.5),
  beta[contact] ~ dnorm(0,0.5),
  phi <- omega + 2,
  omega ~ dexp(1)),
  data = d_in, chains = 4, cores = 4, log_lik = T)
```

```{r}
post = extract.samples(m2)
post$diff_beta_pop = exp(post$beta[,2]) - exp(post$beta[,1])
post$diff_alpha_pop = exp(post$alpha[,2]) - exp(post$alpha[,1])
psis= PSIS(m2, pointwise = T)
psis$culture = d$culture
```

-   so here, note that though there are high K pareto values, they are not as
    high as in the other model (\>1) but still high

```{r}
psis;precis(post)
```

-   now only 3 higher than 0.5 tonga and hawaii (HI low contact, large pop, many
    tools) & (tonga high pop low contact, low tools); yap was also an outlier in
    the previous model
-   It seems that in general, high contact civilizations have more tools,
    irrespective of population size, shown by the huge difference in alphas
-   however, it is now obvious that there is very little real difference between
    high and low contact when considering population size, there is a chance
    that higher contact societies may have slightly less tools than lower
    contact ones when controlling for population but this may be caused by the
    Hawaii outlier, huge pop, no contact, relatively high number of tools

Lets show the estimates on the scale of the data now

```{r}
set.seed(2022)
Kline$log_pop_std = (log(Kline$population) - mean(log(Kline$population)))/sd(log(Kline$population))
fake_n = 1000
log_pop_std = seq(from = -1.5, to = 3, length.out = fake_n)
fake_low = data.frame(pop = log_pop_std, contact = 1)
fake_high = data.frame(pop = log_pop_std, contact = 2)
sim_low = link(m2, data = fake_low)
sim_high = link(m2, data = fake_high)
mu_low = apply(sim_low$lambda, 2, mean)
ci_low = apply(sim_low$lambda, 2, PI, 0.95)
mu_high = apply(sim_high$lambda, 2, mean)
ci_high = apply(sim_high$lambda, 2, PI, 0.95)

p_in = data.frame(
    log_pop_std = log_pop_std,
    mu_low = mu_low,
    ci_low_l = ci_low[1,],
    ci_low_u = ci_low[2,],
    mu_high = mu_high,
    ci_high_l = ci_high[1,],
    ci_high_u = ci_high[2,]) |>
  mutate(
    log_pop = log_pop_std*1.53 + 9,
    pop = exp(log_pop))
```

```{r}
ggplot(data = p_in, aes(x = log_pop_std, y = mu_high)) +
  geom_line(color = "steelblue", alpha = 1.5) +
  geom_ribbon(inherit.aes = F, color = "steelblue", fill = "steelblue", alpha = 0.1,
    aes(x = log_pop_std, ymin = ci_high_l, ymax = ci_high_u)) +
  geom_line(aes(y = mu_low), color = "darkgreen", linetype = "dashed", alpha = 1.5) + 
  geom_ribbon(inherit.aes = F, color = "darkgreen", fill = "darkgreen", alpha = 0.1,
    aes(x = log_pop_std, ymin = ci_low_l, ymax = ci_low_u)) +
  geom_point(inherit.aes = F, color = "darkred", shape = 21, data = Kline,
             aes(x = log_pop_std, y = total_tools, fill = contact)) +
  scale_fill_manual(values = c("high" = "darkred", "low" = NULL)) +
  geom_text(inherit.aes = F,
            data = Kline, aes(x = log_pop_std, y = total_tools+10, label = culture)) +
  theme_minimal() + 
  labs(x = "STD Log Population", y = "Tool Count", title = "Standardized Log Population", fill = NULL)

ggplot(data = p_in, aes(x = pop, y = mu_high)) +
  geom_line(color = "steelblue") +
  geom_ribbon(inherit.aes = F, color = "steelblue", fill = "steelblue", alpha = 0.1,
    aes(x = pop, ymin = ci_high_l, ymax = ci_high_u)) +
  geom_line(aes(y = mu_low), color = "darkgreen", linetype = "dashed") + 
  geom_ribbon(inherit.aes = F, color = "darkgreen", fill = "darkgreen", alpha = 0.1,
    aes(x = pop, ymin = ci_low_l, ymax = ci_low_u)) +
  geom_point(inherit.aes = F, data = Kline, shape = 21, color = "darkred",
             aes(x = population, y = total_tools, fill = contact)) +
  scale_fill_manual(values = c("high" = "darkred", "low" = NA)) +
  theme_minimal() + 
  coord_cartesian(xlim = c(0,3e5)) +
  labs(x = "Population", y = "Tool Count", title = "Population")
```

-   so, really, not a lot of difference in population that relates to tool count
-   seems like contact may be what matters, lets test that using PSIS and WAIC

```{r}
d_in = list(contact = ifelse(d$contact == "low",1,2),
            tools = d$total_tools)
m21 = ulam(alist(
  tools ~ dgampois(mu = lambda, scale = phi),
  log(lambda) <- alpha[contact],
  alpha[contact] ~ dnorm(3,0.5),
  phi <- omega + 2,
  omega ~ dexp(1)),
  data = d_in, chains = 4, cores = 4, log_lik = T)
```

```{r}
compare(m2,m21, func = PSIS)
compare(m2,m21, func = WAIC)
```

-   according to information criterion while the model containing population has
    more weight, it seems that the standard error of the difference is barely
    within the true difference, so the population model slightly edges this
    other one out

# Zero-Inflated Poisson

-   back to the monks example
-   lets simulate some data

```{r}
set.seed(365)
prob_drink = 0.2 # drinks 20% of the time
rate_work = 1 # produces an average of 1 manuscript per day
n = 365 # 1 year of production
drink = rbinom(n, 1, prob_drink)
y = (1-drink)*rpois(n, rate_work)
simplehist(y);simplehist(drink)
```

-   so, some of the 0 lines come from drinking, others come from not completing
    a manuscript naturally
-   now lets create the zero-inflated poisson model

```{r}
curve(dnorm(x,0,1))
curve(dnorm(log(x),-2,1))
```

```{r}
m3 = ulam(alist(
  y ~ dzipois(p, lambda),
  logit(p) <- alpha_p,
  log(lambda) <- alpha_lambda,
  alpha_p ~ dnorm(-1.5,1), # nudged to move prob mass below 0.5
  alpha_lambda ~ dnorm(1,0.5)
), data = list(y = y), chains = 4, cores = 4, log_lik = T)
precis(m3)
```

```{r}
post = extract.samples(m3)
post$nat_alpha_p = inv_logit(post$alpha_p)
post$nat_alpha_lambda = exp(post$alpha_lambda)
precis(post)
```

-   so one thing to note is that p and lambda are untransformed in the output
-   know this because they are equal to alpha lambda and alpha p
-   depending on seed it is easier for me to recover parameters

# Ordinal Categorical Outcomes

-   ordered categories
-   uses a trolley problem data set, I hate trolley problems

```{r}
data("Trolley")
d = Trolley
str(d)
```

```{r}
simplehist(d$response, xlab = "Response")
```

-   response is how appropriate the participant thinks that action is
    -   so it seems like most people didn't think it was appropriate or
        inappropriate

Now to calculate proportionality and log odds

```{r}
prop_k = table(d$response)/nrow(d)
cum_prop_k = cumsum(prop_k)
plot( 1:7 , cum_prop_k , type="b" , xlab="response" ,
ylab="cumulative proportion" , ylim=c(0,1) )
```

-   above is the cumulative proportion of proportion k, below is logit

```{r}
logit_fun = function(x){log(x/(1-x))} # same as built in logit() function
log_cum_prop_k = logit_fun(cum_prop_k)
log_cum_prop_k
plot( 1:7 , log_cum_prop_k , type="b" , xlab="response" ,
ylab="cumulative proportion" , ylim=c(-2,2) )
```

-   so the max odds is infinity or NaN

Now lets build a model where I am going to get the model to estimate the above
cumulative-log-odds for each ordinal category

```{r}
m4 = ulam(alist(
    y ~ dordlogit(0, cutpoints), # 0 is where the linear model phi would normally go
    cutpoints ~ dnorm(0,1.5)), 
    data = list(y = d$response), chains = 4, cores = 4)
```

-   he uses the term cutpoints, I would prefer alpahs, but thats what he uses
-   also this is causing a lot of errros, but, I guess we can just ignore them
    -   seems like it took a while to identify the correct parameters, maybe
        didn't do a good job specifying that prior
-   and this takes significantly longer than any model run so far

```{r}
post = extract.samples(m4)
precis(post, depth = 2);log_cum_prop_k
p_in = data.frame(y = c(log_cum_prop_k, c(apply(post$cutpoints,2,mean),NaN)),
                  x = rep(1:7,2),
                  type = c(rep("proportion",7),rep("estimated",7)) )
ggplot(data = p_in, aes(x = x, y = y, color = type)) + geom_point(alpha = 0.5)
```

-   samples worked great though
-   as you can see, the estimated alpha is almost exactly what was calculated
-   so close that the points are indistinguishable on the graph

```{r}
round(inv_logit(coef(m4)), 3); cum_prop_k
```

-   these are almost exactly the cumulative probabilities
-   now lets look at the outcome probability

```{r}
pk1 = round(dordlogit(1:7,0,coef(m4)),2)
pk1
```

-   so these are the base outcome probabilities of each level of response or y
-   when you subtract 0.5 from them, the probabilities on the higher scales get
    larger while the probabilities on the lower end of response get smaller

```{r}
pk2 = round(dordlogit(1:7,0,coef(m4)-1),2)
pk2
```

```{r}
pk1 - pk2
```

-   so the lowest level probability decreased by 0.08 while the highest
    increased by 0.17

```{r}
sum(pk1*(1:7))
```

-   this implies that the average response is between 4 or 5 but more 4
    responses

```{r}
sum(pk2*(1:7))
```

-   subtracting 1 makes the average response between 5 or 6, huge change towards
    the top end of the scale
-   and if the beta was negative, for instance(-1) then subtracting beta would
    shift it towards smaller probabilites

```{r}
pk3 = round(dordlogit(1:7,0,coef(m4)-(-1)),2)
pk3;pk1-pk3;sum(pk3*(1:7))
```

-   so now the response grew a lot in the lower ends of the scale and decreased
    at the higher ends
-   leading to the average response now being 3 or 4

Now lets add a linear model in here!

-   first lets look at the data

```{r}
d
```

-   case here is just each individual
-   the response is how morally permissable they felt the action was (1 least, 7
    most)
-   action, intention, contact describe what happens in the story
    -   action is whether the harm is caused by something someone did
    -   intention is whether someone knew the harm their action would cause
    -   contact is whether someone actually "pushed" a person instead of just
        doing something that indirectly causes harm
        -   theoretically contact always involved action, but that is not the
            way this is coded

Linear Model

-   below is a way to model interaction
-   so essentially, BetaI contains all possible interactions withinit
-   so if A and C are 0, then betaA, betaIA, betaC,betaIC all become 0 leaving
    only betaI
-   if I is 0 then it cancels out that whole second line
-   however, if either one of A, C or both are not 0 and I is not 0 then I
    multiplies that whole term set automatically creating interactions

```{r}
d_in = list(R = d$response, A = d$action, I = d$intention, C = d$contact)
m4 = ulam(alist(
    R ~ dordlogit(phi, alphas),
    phi <- betaA*A + betaC*C + BetaI*I, # doing something weird to create interaction
    BetaI <- betaI + betaIA*A + betaIC*C,
    c(betaA,betaC,betaI,betaIA,betaIC) ~ dnorm(0,0.5),
    alphas ~ dnorm(0,1.5)), 
    data = d_in, chains = 4, cores = 4)
```

```{r}
precis(m4)
plot(precis(m4))
```

-   so, the presence of all the story features make the moral rating worse
-   interestingly, the interaction of intention (knowing the harm of the action)
    and contact (physically touching the harmed person) decreases the morality
    far more than anything else
-   he generates a bunch of posterior graphs but I am not super interested in
    that for this context
-   but basically he shows how the posterior reflects the bad interaction
    between intention and contact and decreases the morality rating of the
    action

# Ordinal Categorical Predictors

-   so we conquered ordinal outcomes, what about ordinal predictors?
-   he introduces the dirichlet distribution here

```{r}
library(gtools)
delta2 = rdirichlet(10, alpha = rep(2,7))
str(delta2)
```

-   so output is 10 samples (10 rows) of a vector of 7 values
-   each value is probability for the particular vector, based on a seed of 2

```{r}
h = 4
plot( NULL , xlim=c(1,7) , ylim=c(0,0.4) , xlab="index" , ylab="probability" )
for (i in 1:nrow(delta2)) lines(1:7, delta[i,], type="b",
pch=ifelse(i==h,16,1), lwd=ifelse(i==h,4,1.5),
col=ifelse(i==h,"blue",col.alpha("blue",0.7)))
```

-   the probabilities within each vector can vary pretty wildly
-   this is good, it allows the "distance" between each lil delta to vary wildly
-   lets look at other vectors of dirichlet distribution

```{r}
delta = rdirichlet(10, alpha = rep(0.1,7))
str(delta)
h = 4
plot( NULL , xlim=c(1,7) , ylim=c(0,1) , xlab="index" , ylab="probability" )
for (i in 1:nrow(delta)) lines(1:7, delta[i,], type="b",
pch=ifelse(i==h,16,1), lwd=ifelse(i==h,4,1.5),
col=ifelse(i==h,"blue",col.alpha("blue",0.7)))
```

-   so when alpha is \<1, the distribution creates more "edge cases" or sticks
    probability density in more specific spots, while giving none to others
    -   in this case, that is like saying some categories don't matter at all
        while others matter a lot
-   when alpha \> 1, the probability density is more even
-   this is expected because mean = component/sum of all components
-   variance = complicated formula, but as the vector grows longer, increase in
    number of components, variance shrinks
-   Marginals =

```{r}
n = 10
delta = rdirichlet(10, alpha = rep(0.5,n))
#str(delta)
for(i in 1:nrow(delta)){
  print(mean(delta[,i]))
}
h = 4
plot( NULL , xlim=c(1,n) , ylim=c(0,1) , xlab="index" , ylab="probability" )
for (i in 1:nrow(delta)) lines(1:n, delta[i,], type="b",
pch=ifelse(i==h,16,1), lwd=ifelse(i==h,4,1.5),
col=ifelse(i==h,"blue",col.alpha("blue",0.7)))
```

Going to include education in model but need to fix

```{r}
for(i in levels(d$edu)){print(i)}
level_vec = c(6,1,8,4,7,2,5,3)
d$edu2 = level_vec[d$edu] # integer value for each level
for(i in 1:length(levels(d$edu))){print(paste0(levels(d$edu)[i], ": ", level_vec[i]))}
```

```{r}
d_in = list(R = d$response, 
            A = d$action, 
            I = d$intention, 
            C = d$contact,
            E = d$edu2,
            delta_prior = rep(2,7)) # the alpha prior vector for dirichlet distribution
m4 = ulam(alist(
    R ~ dordlogit(phi, alphas),
    phi <- betaA*A + betaC*C + BetaI*I + 
      betaE*sum(delta_j[1:E]), # this is the new ordinal beta, betaE X sum of ordinal categorical like notes
    BetaI <- betaI + betaIA*A + betaIC*C,
    c(betaA,betaC,betaI,betaIA,betaIC, betaE) ~ dnorm(0,0.5),
    alphas ~ dnorm(0,1.5),
    vector[8]: delta_j <<- append_row(0,delta), # adding that 0 for the first education level, since the last level is "removed"
    simplex[7]: delta ~ ddirichlet(delta_prior)
    ), 
    data = d_in, chains = 4, cores = 4)
```

-   took about 7 minutes to sample

```{r}
precis(m4, depth = 2)
```

-   so the difference between some college and bachelors degree is very small if
    I am interpreting these delta parameters correctly
