---
title: "Chapter 11"
author: "P Winston Miller"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rethinking)
library(dagitty)
library(ggdag)
library(patchwork)
library(paletteer)
set_cmdstan_path("/Users/winston/miniconda3/envs/stan/bin/cmdstan")
```

# Logistic Regression

```{r}
data("chimpanzees")
d = chimpanzees
```

```{r}
d$treatment = 1 + d$prosoc_left + 2*d$condition
xtabs( ~ treatment + prosoc_left + condition , d )
```

-   condition 0 = no partner
-   condition 1 = partner
-   prosocial left = if prosocial option was on on left
    -   1 = yes
    -   2 = no
-   treatment 1 = prosocial option right, no partner
-   treatment 2 = prosocial option left, no partner
-   treatment 3 = prosocial option right, partner
-   treatment 4 = prosocial option left, partner

Basic Model - just an intercept for treatment - going to build several of these
with different priors for alpha

```{r}
m1_in = list(pulled_left = d$pulled_left)
m1_flat = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha,
        alpha ~ dnorm(0,10)),
        data = m1_in, chains = 4, cores = 4)

m1_sharp = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha,
        alpha ~ dnorm(0,2)),
        data = m1_in, chains = 4, cores = 4)

m1_sharper1 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha,
        alpha ~ dnorm(0,1.5)),
        data = m1_in, chains = 4, cores = 4)

m1_sharper2 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha,
        alpha ~ dnorm(0,1)),
        data = m1_in, chains = 4, cores = 4)

m1_sharpest1 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha,
        alpha ~ dnorm(0,0.5)),
        data = m1_in, chains = 4, cores = 4)
m1_sharpest2 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha,
        alpha ~ dnorm(0,0.1)),
        data = m1_in, chains = 4, cores = 4)
```

```{r}
prior_flat = extract.prior(m1_flat)
prior_sharp = extract.prior(m1_sharp)
prior_sharper1 = extract.prior(m1_sharper1)
prior_sharper2 = extract.prior(m1_sharper2)
prior_sharpest1 = extract.prior(m1_sharpest1)
prior_sharpest2 = extract.prior(m1_sharpest2)
```

```{r}
prior_in = data.frame(
  prior_samples = c(
    inv_logit(prior_flat$alpha),
    inv_logit(prior_sharp$alpha),
    inv_logit(prior_sharper1$alpha),
    inv_logit(prior_sharper2$alpha),
    inv_logit(prior_sharpest1$alpha),
    inv_logit(prior_sharpest2$alpha)),
  prior_shape = c(
    rep("flat - N(0,10)",1000),
    rep("sharp - N(0,2)",1000),
    rep("sharper - N(0,1.5)",1000),
    rep("sharper - N(0,1)",1000),
    rep("sharpest - N(0, 0.5)",1000),
    rep("sharpest - N(0, 0.1)",1000)))

ggplot(data = prior_in, aes(x = prior_samples, color = prior_shape)) +
  geom_density()+ 
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  theme_minimal() +
  theme(angle = 45) +
  labs(x = "Probablity Puleld Left", y = "Probability Density",
       color =  "Prior Shape") +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
```

-   so here, a traditionally flat prior pulls probability towards 0, 1 so,
    monkey never pulls left or always pulls left
-   as the distributions get sharper, they flatten out, N(0,1.5) is about as
    flat as can be
-   this holds true until normal prior goes below one, then it starts to really
    concentrate around 50%
-   N(0,0.1) makes the model VERY sure that pulls are basically random

Now, lets do the same thing for Beta

Beta Prior Test Model - making the alpha relatively flat because I don't know
the chimps preference for pulling handles, it might change for any reason

```{r}
m2_in = list(pulled_left = d$pulled_left, treatment = d$treatment)
m2_flat = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha + beta[treatment],
        alpha ~ dnorm(0,1.5),
        beta[treatment] ~ dnorm(0,10)),
        data = m2_in, chains = 4, cores = 4)

m2_sharp = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha + beta[treatment],
        alpha ~ dnorm(0,1.5),
        beta[treatment] ~ dnorm(0,2)),
        data = m2_in, chains = 4, cores = 4)

m2_sharper1 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha + beta[treatment],
        alpha ~ dnorm(0,1.5),
        beta[treatment] ~ dnorm(0,1.5)),
        data = m2_in, chains = 4, cores = 4)

m2_sharper2 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha + beta[treatment],
        alpha ~ dnorm(0,1.5),
        beta[treatment] ~ dnorm(0,1)),
        data = m2_in, chains = 4, cores = 4)

m2_sharpest1 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha + beta[treatment],
        alpha ~ dnorm(0,1.5),
        beta[treatment] ~ dnorm(0,0.5)),
        data = m2_in, chains = 4, cores = 4)
m2_sharpest2 = ulam(
  alist(pulled_left ~ dbinom(1,p),
        logit(p) <- alpha + beta[treatment],
        alpha ~ dnorm(0,1.5),
        beta[treatment] ~ dnorm(0,0.1)),
        data = m2_in, chains = 4, cores = 4)
```

```{r}
prior_flat = extract.prior(m2_flat)
prior_sharp = extract.prior(m2_sharp)
prior_sharper1 = extract.prior(m2_sharper1)
prior_sharper2 = extract.prior(m2_sharper2)
prior_sharpest1 = extract.prior(m2_sharpest1)
prior_sharpest2 = extract.prior(m2_sharpest2)
```

```{r}
pflat = sapply(1:4, function(k) inv_logit(prior_flat$alpha + prior_flat$beta[,k]))
head(pflat)
```

-   above code produces 4 columns
-   to get the probability of pulling left, and how sociability affects pulling
    left
-   need to add alpha + the beta for each treatment together, then convert to
    probability by inverting the logit
-   I just want to look at the difference between the first two treatments,
    which is no partner, prosocial option on left (t1) prosocial on right (t2)
-   basically, this is going to tell me the prior difference the model assumes
    chimps have for just pulling the left lever for no reason

```{r}
psharp = sapply(1:4, function(k) inv_logit(prior_sharp$alpha + prior_sharp$beta[,k]))
psharper1 = sapply(1:4, function(k) inv_logit(prior_sharper1$alpha + prior_sharper1$beta[,k]))
psharper2 = sapply(1:4, function(k) inv_logit(prior_sharper2$alpha + prior_sharper2$beta[,k]))
psharpest1 = sapply(1:4, function(k) inv_logit(prior_sharpest1$alpha + prior_sharpest1$beta[,k]))
psharpest2 = sapply(1:4, function(k) inv_logit(prior_sharpest2$alpha + prior_sharpest2$beta[,k]))
prior_in = data.frame(
  prior_samples = c(
    abs(pflat[,1] - pflat[,2]),
    abs(psharp[,1] - psharp[,2]),
    abs(psharper1[,1] - psharper1[,2]),
    abs(psharper2[,1] - psharper2[,2]),
    abs(psharpest1[,1] - psharpest1[,2]),
    abs(psharpest2[,1] - psharpest2[,2])),
  prior_shape = c(
    rep("flat - N(0,10)",1000),
    rep("sharp - N(0,2)",1000),
    rep("sharper - N(0,1.5)",1000),
    rep("sharper - N(0,1)",1000),
    rep("sharpest - N(0, 0.5)",1000),
    rep("sharpest - N(0, 0.1)",1000)))
```

```{r}
ggplot(data = prior_in, aes(x = prior_samples, color = prior_shape)) +
  geom_density()+ 
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  theme_minimal() +
  theme(angle = 45) +
  labs(x = "Probablity Puleld Left", y = "Probability Density",
       color =  "Prior Shape") +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
```

-   okay, so once again the sharpest prior basically says the difference between
    treatments is nothing
-   lets remove that so I can actually see the diffferences in the others

```{r}
me1 = round(mean(abs(pflat[,1] - pflat[,2])),3)
me2 = round(mean(abs(psharp[,1] - psharp[,2])),3)
me3 = round(mean(abs(psharper1[,1] - psharper1[,2])),3)
me4 = round(mean(abs(psharper2[,1] - psharper2[,2])),3)
me5 = round(mean(abs(psharpest1[,1] - psharpest1[,2])),3)
prior_in = data.frame(
  prior_samples = c(
    abs(pflat[,1] - pflat[,2]),
    abs(psharp[,1] - psharp[,2]),
    abs(psharper1[,1] - psharper1[,2]),
    abs(psharper2[,1] - psharper2[,2]),
    abs(psharpest1[,1] - psharpest1[,2])),
  prior_shape = c(
    rep(paste0("flat - N(0,10) - ", me1),1000),
    rep(paste0("sharp - N(0,2) - ", me2),1000),
    rep(paste0("sharper - N(0,1.5) - ", me3),1000),
    rep(paste0("sharper - N(0,1) - ", me4),1000),
    rep(paste0("sharpest - N(0, 0.5) - ", me5),1000)))

ggplot(data = prior_in, aes(x = prior_samples, color = prior_shape)) +
  geom_density()+ 
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  theme_minimal() +
  theme(angle = 45) +
  labs(x = "Probablity Puleld Left", y = "Probability Density",
       color =  "Prior Shape") +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
```

-   so, similar to alpha, the more "flat" the prior is, the more it concentrates
    probability to the treatment has no effect (0%) or treatment completely
    control behavior (100%)
-   as priors get sharper, probability gets more concentrated towards treatment
    having no effect, with a degrading likelihood of effect as effect gets
    stronger
-   means show the difference in probability for effect of treatment on pulling
    the left lever
    -   remember, just looking at the contrast between 2 foods on right or 2
        foods on left
    -   outcome is pulling left lever
-   large differences will shine through, but priors need to be skeptical of
    those to reduce overfitting
-   so now that I have decided appropriately skeptical priors, need to build the
    full model

Full Model - I am including an intercept for each chimpanzee - he doesn't say
this, but this seems like a typical mixed model for time series - where each
chimpanzee pulls the lever x number of times and has its own random effect

```{r}
set.seed(1992)
m3_in = list(
  pulled_left = d$pulled_left,
  treatment = d$treatment,
  actor = d$actor)
m3 = ulam(alist(
  pulled_left ~ dbinom(1,p),
  logit(p) <- alpha[actor] + beta[treatment],
  alpha[actor] ~ dnorm(0,1.5),
  beta[treatment] ~ dnorm(0,0.5)),
  data = m3_in, chains = 4, cores = 4, log_lik = T)
```

```{r}
precis(m3, depth = 2)
```

-   so each chimp has its own intercept, indicating its own log-odds to pull the
    left lever
-   so seems like chimp 2 almost always pulls the left lever, log-odds above 4
    is basically always 1, log-odds below -4 is basically always 0

First, lets see the probability that any given chimpanzee has a lever preference

```{r}
post = extract.samples(m3)
p_left = data.frame(inv_logit(post$alpha)) |>
  pivot_longer(cols = everything(),
               names_to = "Chimps",
               values_to = "Prob Pulled Left") |> as.data.frame()
  
ggplot(data = p_left, aes(x = `Prob Pulled Left`, y = Chimps, color = Chimps)) +
  geom_line(linewidth = 0.8) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 21, size = 3) +
  scale_x_continuous(labels = scales::percent, breaks = seq(0,1,0.1)) + 
  theme_minimal() +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
```

-   so clearly most chimps have a mean preference for the right lever
-   chimps 2 and 7 have a mean preference for the left
-   in fact, it looks like chimp 2 pretty much never pulled the right lever

Since the model accounts for these differences, I can now look at log-odds for
the treatments

```{r}
treatment_affect = data.frame(post$beta) |>
  rename("R/NP" = "X1",
         "L/NP" = "X2",
         "R/P" = "X3",
         "L/P" = "X4") |>
  pivot_longer(cols = everything(),
               names_to = "Treatment",
               values_to = "LogOdds") |> as.data.frame()
  
ggplot(data = treatment_affect, aes(x = LogOdds, y = Treatment, color = Treatment)) +
  geom_line(linewidth = 0.8) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 21, size = 3) +
  scale_x_continuous(breaks = seq(-2,2,0.25)) + 
  theme_minimal() +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
```

-   0 is no effect,
-   the treatment doesn't seem to have an effect on the chimps
-   next, would want to look at the contrasts, so how does there being a partner
    change the pull?
    -   doesn't appear to be much but need to check

```{r}
treatment_contrast = data.frame(
  LogOdds = c(
    c(post$beta[,1] - post$beta[,3]),
    c(post$beta[,2] - post$beta[,4])),
  Hand = c(
    rep("Right",2000),
    rep("Left",2000)))
  
ggplot(data = treatment_contrast, aes(x = LogOdds, y = Hand, color = Hand)) +
  geom_line(linewidth = 0.8) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 21, size = 3) +
  scale_x_continuous(breaks = seq(-2,2,0.25)) + 
  theme_minimal() +
  scale_colour_paletteer_d("ggsci::category20_d3",direction = 1)
```

-   so, I think Dick's interpretation in the book is incorrect, the more
    negative something is in this contrast, the more pulls happened when partner
    is present
-   so, looks like there were slightly more left-handed pulls when partner was
    present than without since it is slightly more negative
-   but, there is clearly no effect of partner present/absent having an effect

Now, check the posterior predictions 1. Calculate the probability that each
chimp pulled lef

```{r}
pulls_og = unclass(by(d$pulled_left, list(d$actor, d$treatment), mean))
attr(pulls_og, "call") = NULL
pulls_og = as.data.frame(pulls_og)
```

-   columns are treatments
-   rows are actors
-   values are proportions of left pulls

```{r}
sim_post = link(m3, data = list(actor=rep(1:7,each=4) , treatment=rep(1:4,times=7)))
sim_mu = apply(sim_post,2,mean)
sim_mu = as.data.frame(matrix(sim_mu, ncol = 4, nrow = 7, byrow = T))
colnames(sim_mu) = c(1,2,3,4)
```

```{r}
sim_mu - pulls_og
apply(sim_mu - pulls_og,2,mean)
```

-   on average model is off by less than 2% on its posterior predictions, which
    is pretty good

# Poisson Regression

-   this doing the whole oceanic islands tool thing
-   basically, there is an unknown maximum of tools, and number of tools is a
    discrete count
-   much like in logistic regression, the transformation has a big effect on the
    priors
-   transformation is log, so if prior is normal, it is then changed to
    log-normal for outcome

Simulate Typical Priors

```{r}
curve( dlnorm( x , 0 , 10 ) , from=0 , to=100 , n=200, col = "black")
curve( dlnorm( x , 3 , 0.5 ) , from=0 , to=100 , n=200, add = T, col = "red")
curve( dlnorm( x , 3 , 1 ) , from=0 , to=100 , n=200, add = T, col = "blue")
curve( dlnorm( x , 10 , 20 ) , from=0 , to=100 , n=200, add = T, col = "green")
```

-   so, need to do run this especially for log-normal outcomes

```{r}
set.seed(10)
n = 100
alpha =  rnorm(n, 3, 0.5 )
beta = rnorm(n, 0, 10 ) # 1
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}

beta = rnorm(n, 0, 5 ) # 2
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}

beta = rnorm(n, 0, 2 ) # 3
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}

beta = rnorm(n, 0, 1 ) # 4
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}

beta = rnorm(n, 0, 0.5 )
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}

beta = rnorm(n, 0, 0.25 )
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}

beta = rnorm(n, 0, 0.1 )
plot(NULL, xlim=c(-2,2), ylim=c(0,100) )
for (i in 1:n){
  curve(exp(alpha[i] + beta[i]*x) , add=TRUE , col=grau() )}
```

this is assuming standardization

-   note that all the lines still go through 0, which is what I want since this
    is standardized

1.  Beta 1: means that the model says the number of tools collapses once the
    population decreases from mean value, and explodes once population exceeds
    mean value
    -   or vice versa.
    -   this makes no sense, and so is bad prior
2.  Beta 2: similar, but not as extreme due to not as many negative values
    -   I think it would be possible to force this to be positive, but, probably
        not necessary

<!-- -->

3.  Beta 3: So as the standard deviation gets smaller, the swings get less
    extreme

4.  Beta 5: These lines are almost completely flat, model assumes no effect of
    populationn size at all

    -   this is also not good, I am pretty sure population size matters

5.  Beta 4: These lines still seem to have too much swing

6.  Beta 5: This seems like the best, the model assumes only large differences
    in standard log population are going to affect the number of tools which
    seems right

Now to Graph on the real values

```{r}
set.seed(10)
n = 100
alpha =  rnorm(n, 3, 0.5 )
beta = rnorm(n, 0, 0.25 )
x_seq = seq(from=log(100) , to=log(200000) , length.out=100 )
lambda = sapply( x_seq , function(x) exp( alpha + beta*x))
plot( NULL , xlim=range(x_seq) , ylim=c(0,500) , xlab="log population" ,
ylab="total tools" )
for ( i in 1:n ){lines(x_seq, lambda[i,], col=grau(), lwd=1.5)}

plot( NULL , xlim=range(exp(x_seq)) , ylim=c(0,500), xlab="population",
ylab="total tools" )
for ( i in 1:n ) lines( exp(x_seq) , lambda[i,] , col=grau() , lwd=1.5 )
```

```{r}
data("Kline")
d_in = list(tools = Kline$total_tools,
            log_pop_std = scale(log(Kline$population)),
            contact = ifelse(Kline$contact == "low",1,2))
```

```{r}
m41 = ulam(alist(
  tools ~ dpois(lambda),
  log(lambda) <- alpha,
  alpha ~ dnorm(3,0.5)),
  data = d_in, chains = 4, cores = 4, log_lik = T)

m42 = ulam(alist(
  tools ~ dpois(lambda),
  log(lambda) <- alpha[contact] + beta[contact]*log_pop_std,
  alpha[contact] ~ dnorm(3,0.5),
  beta[contact] ~ dnorm(0,0.25)),
  data = d_in, chains = 4, cores = 4, log_lik = T)
```

```{r}
precis(m41, depth = 2);precis(m42, depth = 2);compare(m41,m42,func="PSIS")
```

-   highly influential points (expected due to small sample size)
-   m42 is clearly better for a lot of reasons
-   lets see what those points are

```{r}
k = PSIS(m42, pointwise = T)
row.names(k) = Kline$culture
k[order(k$k, decreasing = T),]
```

-   so,Tonga, Hawaii, and Trobriand are all highly influential

```{r}
Kline$log_pop_std = (log(Kline$population) - mean(log(Kline$population)))/sd(log(Kline$population))
Kline
```

-   reason why Hawaii is influential is obvious, very high population, very high
    tools, but low contact, which is very different than the others
-   Tonga is a much higher population than everybody else as well
-   Yap has a very low population, but a lot of tools
-   Trobrians has the opposite problem, relatvely high pop, high contact and
    like no tools
-   okay, now need to calculate posterior predictions and plot them over the
    data points to see how that influences the model

```{r}
set.seed(2022)
fake_n = 1000
log_pop_std = seq(from = -1.5, to = 3, length.out = fake_n)
fake_low = data.frame(log_pop_std = log_pop_std, contact = 1)
fake_high = data.frame(log_pop_std = log_pop_std, contact = 2)
sim_low = link(m42, data = fake_low)
sim_high = link(m42, data = fake_high)
mu_low = apply(sim_low, 2, mean)
ci_low = apply(sim_low, 2, PI, 0.95)
mu_high = apply(sim_high, 2, mean)
ci_high = apply(sim_high, 2, PI, 0.95)
```

```{r}
p_in = data.frame(
    log_pop_std = log_pop_std,
    mu_low = mu_low,
    ci_low_l = ci_low[1,],
    ci_low_u = ci_low[2,],
    mu_high = mu_high,
    ci_high_l = ci_high[1,],
    ci_high_u = ci_high[2,]) |>
  mutate(
    log_pop = log_pop_std*1.53 + 9,
    pop = exp(log_pop))
```

```{r}
ggplot(data = p_in, aes(x = log_pop_std, y = mu_high)) +
  geom_line(color = "steelblue") +
  geom_ribbon(inherit.aes = F, color = "steelblue", fill = "steelblue", alpha = 0.1,
    aes(x = log_pop_std, ymin = ci_high_l, ymax = ci_high_u)) +
  geom_line(aes(y = mu_low), color = "darkgreen", linetype = "dashed") + 
  geom_ribbon(inherit.aes = F, color = "darkgreen", fill = "darkgreen", alpha = 0.1,
    aes(x = log_pop_std, ymin = ci_low_l, ymax = ci_low_u)) +
  geom_point(inherit.aes = F, color = "darkred", shape = 21, data = Kline,
             aes(x = log_pop_std, y = total_tools, fill = contact)) +
  scale_fill_manual(values = c("high" = "darkred", "low" = NULL)) +
  geom_text(inherit.aes = F,
            data = Kline, aes(x = log_pop_std, y = total_tools+10, label = culture)) +
  theme_minimal() + 
  labs(x = "STD Log Population", y = "Tool Count", title = "Standardized Log Population", fill = NULL)
```

-   so basically Hawaii is pretty much controlling the direction of the
    low-contact\*population line
-   and the model is not fitting Tonga well

```{r}
ggplot(data = p_in, aes(x = pop, y = mu_high)) +
  geom_line(color = "steelblue") +
  geom_ribbon(inherit.aes = F, color = "steelblue", fill = "steelblue", alpha = 0.1,
    aes(x = pop, ymin = ci_high_l, ymax = ci_high_u)) +
  geom_line(aes(y = mu_low), color = "darkgreen", linetype = "dashed") + 
  geom_ribbon(inherit.aes = F, color = "darkgreen", fill = "darkgreen", alpha = 0.1,
    aes(x = pop, ymin = ci_low_l, ymax = ci_low_u)) +
  geom_point(inherit.aes = F, data = Kline, shape = 21, color = "darkred",
             aes(x = population, y = total_tools, fill = contact)) +
  scale_fill_manual(values = c("high" = "darkred", "low" = NA)) +
  theme_minimal() + 
  labs(x = "Log Population", y = "Tool Count", title = "Standardized Log Population")
```
